---
title: "Viz and Share"
output:
  html_document:
    df_print: paged
---

# The data cleaned and in the right format?

As soon as you opena data file, you must check the data types with **str()**:

```{r, eval=TRUE, echo=TRUE, message=FALSE}
rm(list = ls())

###
git='https://github.com/DACSS-Fundamentals/'
org='vizandshare/raw/main/'
file='ciadata.rds'
gitLocation=paste0(git,org)
fullLocation=paste0(gitLocation,file)
cia=readRDS(url(fullLocation))

##
str(cia)
```


# You may have several variables...but

You just do not open a data file for a random reason. You have a purpose. 

## Which is your variable of interest?

* Life Expectancy

## Which are you considering as explanatory variables?

* Infant mortality rate
* Maternal mortality rate

## Any Control Variable?

* GINI index
* Region

# Exploration

Do not plan any advanced work if you have not explored the set of columns you need.

## Univariate : Your variable of interest

### key stats

```{r}
summary(cia$life_exp)
```

The above may be supplemented by:
```{r}
library(moments)
skewness(cia$life_exp)
```

The tail is to the left (would be to the right if positive)

### Visual

How to highlight the stats above?

```{r}
# call the library
library(ggplot2) # do not expect an important message
```

```{r}
# tell ggplot the data you will use 
base=ggplot(data=cia) # I call it 'base'
```

```{r}
# decide the plot for the variable in the data
# and see it
base + geom_boxplot(aes(x=life_exp))

```
```{r}
# this is another alternative
base + geom_histogram(aes(x=life_exp),bins = 10)
```

Let's choose the boxplot and add some **textual** decorations:


```{r}
# detecting the worst case
worstCase=cia[which.min(cia$life_exp),'name']
# preparing a message
theCaption=paste("Source:CIA\n","Worst case:",worstCase)  

# text for titles
theTitle='People live more than 70 years around the world'
theSubtitle="147 countries, 2023"

# time te REPLOT
box=base + geom_boxplot(aes(x=life_exp))
box + labs(title=theTitle,
           subtitle = theSubtitle,
           caption = theCaption,
           x= 'life expectancy')
```

## Bivariate

### Key stats

- what about a correlation
```{r, message=FALSE}
corrtable::correlation_matrix(df = cia[,-c(1,5,6)],
                              use = 'lower',
                              replace_diagonal = T)

```

### Visuals

- Visualizing the correlation

```{r}
library(ggcorrplot) # may need installation

# 1. Compute the correlation matrix
corr_matrix <- cor(cia[, c('life_exp',"infant", "maternal")])

# 2. Plot
ggcorrplot(corr_matrix)
```

# Time to test your Hypothesis!

## Key stat

```{r}
# the H
hypo=formula(life_exp ~ infant + maternal + gini)
```

The test

```{r}
model <- lm(hypo, data = cia)

#then
summary(model)
```

** Above, you just found some important difference between correlation and regression.**

## Visual

```{r}
library(sjPlot)
plot_models(model)
```


# playing with simple commands and find some insight?

If this is the data:

```{r}
head(cia)
```
## Life expectancy by region?


```{r}
# version 1
base + geom_boxplot(aes(x=life_exp,y=region))
```
```{r}
# version 2
base + geom_boxplot(aes(x=life_exp,y=reorder(region,life_exp,median)))

```

Which of the above gave you something else to state?


## A box plot for each var?


Remember this is not right:

```{r}
base+geom_boxplot(aes(x=infant,y=infant)) +
     geom_boxplot(aes(x=maternal,y=maternal)) + labs(y='')

```

This is a wide shape:


```{r}
head(cia[,c(1,2,3,4)])
```

We need a long shape for **ggplot2**:

```{r}
ciaLong=tidyr::pivot_longer(data=cia[,c(1,2,3,4)],
                                cols=!name, 
                                names_to = "vars",
                                values_to = "index") 
ciaLong
```

```{r}
# then
base = ggplot(data=ciaLong)
base + geom_boxplot(aes(x=vars,y=index))
```

Good data, but visually still weak. What happened? Each variable has different range (min - max). Then:

**Faceting**

```{r}
base2 = ggplot(data=ciaLong)
base2 + geom_boxplot(aes(y=index)) + 
  facet_wrap(~ vars, scales = "free")
```

Facetting is very useful:

```{r}
base2 = ggplot(data=ciaLong)
base2 + geom_histogram(aes(x=index),bins=10) + 
  facet_wrap(vars~., scales = "free", ncol = 1)
```

## Mortality worst group?

```{r}
head(cia[,c('infant','maternal')])
```
```{r}
# Find the value that marks the top 25% (worst)
inf_bad <- quantile(cia$infant, 0.75, na.rm = TRUE)
mat_bad   <- quantile(cia$maternal, 0.75, na.rm = TRUE)

# '1' if in the worst quartile:
cia$inf_worst <- ifelse(cia$infant >= inf_bad, 1, 0)
cia$mat_worst   <- ifelse(cia$maternal >= mat_bad, 1, 0)

# flag worst of worst
cia$worst_infmat=cia$inf_worst + cia$mat_worst

cia$worst_infmat=ifelse(cia$worst_infmat==2,T,F)
```

Report:
```{r}
aggregate(worst_infmat ~ region, data = cia, FUN = sum, na.rm = TRUE)
```


## Mortality best group?

```{r}
inf_good <- quantile(cia$infant, 0.25, na.rm = TRUE)
mat_good   <- quantile(cia$maternal, 0.25, na.rm = TRUE)

# NOW 'less than'
cia$inf_best <- ifelse(cia$infant <= inf_good, 1, 0)
cia$mat_best   <- ifelse(cia$maternal <= mat_good, 1, 0)

# flag worst of worst
cia$best_infmat=cia$inf_best + cia$mat_best

cia$best_infmat=ifelse(cia$best_infmat==2,T,F)
```

Report:

```{r}
aggregate(best_infmat ~ region, data = cia, FUN = sum, na.rm = TRUE)
```

Plotting!

```{r}
# saving as object
counts_worst <- aggregate(worst_infmat ~ region, data = cia, FUN = sum)
# filtering
counts_worst=counts_worst[counts_worst$worst_infmat>0,]

# plotting
base_worst=ggplot(counts_worst)

base_worst + geom_col(aes(x = reorder(region, -worst_infmat), 
                       y = worst_infmat)) 
```

and....

```{r}
# saving as object
counts_best <- aggregate(best_infmat ~ region, data = cia, FUN = sum)
# filtering
counts_best=counts_best[counts_best$best_infmat>0,]

# plotting
base_best=ggplot(data=counts_best)

base_best + geom_col(aes(x = reorder(region, -best_infmat), 
                       y = best_infmat)) 
```

